Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-23 manuscript/manuscript.tex
began adding notes about limit to manuscript

To test the models on real data, we used actual sequences of influenza HA. 
The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
HA consists of a number of different subtypes.
Sequences within a subtype have $>$75\% amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
Although the nucleotide model used with \texttt{RAxML} to infer this tree topology is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.

@@

\subsection*{pbMutSel inference with \texttt{PhyloBayes-MPI}.}
For \ref{fig:phylobayes}, we fit a pbMutSel model to the full HA tree. 
We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
\skhcomment{re: measure of convergence. I need add some analysis to fully address this point. 1. I am going to run 2 chains and assess the congruency between these two chains. 2. I am going to look at the traces using Andrew Rambaut's program tracer. Both of these are in progress}
We used \texttt{PhyloBayes-MPI} program \texttt{readpb_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences.

@@

\begin{supptable}[t!]
\caption{\label{supptab:fit_params}
{\bf Model parameters fit to a low divergence tree.}
We fit GY94 models and an ExpCM defined by H1 deep mutational scanning preferences to the ``low divergence from H1" tree in \ref{fig:compete}. 
We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay} and simulate the sequences in \ref{fig:simulations}. 
} 
     \begin{tabular}{ccc}
        \hline
         Model & Parameters \\ \hline
         GY94 &  {\shortstack{$\kappa = 3.17$, $\omega = 0.10$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.13$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$,  $\phi_{2G} = 0.19$,\\ $\phi_{3, A} = 0.35$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
         GY94 + $\Gamma\omega$ & {\shortstack{$\alpha_\omega = 0.51$, $\beta_\omega = 3.92$, $\kappa = 3.49$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.14$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$, $\phi_{2,G} = 0.20$, \\$\phi_{3,A} = 0.355$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
         ExpCM(H1) &  {\shortstack{$\beta = 1.22$, $\kappa = 3.24$, $\omega = 0.13$,\\ $\phi_A = 0.37$, $\phi_C = 0.18$, $\phi_G = 0.23$}}\\ \hline
      \end{tabular}
\end{supptable}
\clearpage

@@

>>>>

To test the models on real data, we used actual sequences of influenza HA. 
The topology of HA phylogenetic trees (\ref{fig:empirical_trees}) makes these sequences an interesting test case for branch-length estimation.
HA consists of a number of different subtypes.
Sequences within a subtype have $>$68\% amino-acid identity, but sequences in different subtypes have as little as 38\% identity.
However, HA proteins from all subtypes have a highly conserved structure that performs a highly conserved function~\citep{ha2002h5,russell2004h1}.
We used \texttt{RAxML}~\citep{stamatakis2006raxml} with a nucleotide substitution model  (GTRCAT) to infer a phylogenetic tree for 87 HA sequences drawn from 14 of the 18 subtypes (we excluded bat influenza and three other rare subtypes).
For the rest of this paper, we fix the tree topology to this \texttt{RAxML}-inferred tree.
Although the nucleotide model used with \texttt{RAxML} to infer this tree topology is probably less accurate than codon models, the modular subtype structure of the HA phylogeny means that most of the phylogenetic uncertainty lies in the length of the long branches separating the subtypes rather than in the tree topology itself.

@@

\subsection*{pbMutSel inference with \texttt{PhyloBayes-MPI}.}
For \ref{fig:phylobayes}, we fit a pbMutSel model to the full HA tree. 
We ran one chain for 5000 steps, saved every sample, and discarded the first 500 samples as a burnin. 
\skhcomment{re: measure of convergence. I need add some analysis to fully address this point. 1. I am going to run 2 chains and assess the congruency between these two chains. 2. I am going to look at the traces using Andrew Rambaut's program tracer. Both of these are in progress}
We used \texttt{PhyloBayes-MPI} program \texttt{readpb\_mpi} to compute the majority-rule consensus tree and the posterior average site-specific amino-acid preferences.

@@

\begin{supptable}[t!]
\caption{\label{supptab:fit_params}
{\bf Model parameters fit to a low divergence tree.}
We fit GY94 models and an ExpCM defined by H1 deep mutational scanning preferences to the ``low divergence from H1" tree in \ref{fig:compete}. 
We used these model parameters calculate the expected pairwise sequence identity in \ref{fig:decay} and simulate the sequences in \ref{fig:simulations}. 
\skhcomment{check this numbers}
} 
     \begin{tabular}{ccc}
        \hline
         Model & Parameters \\ \hline
         GY94 &  {\shortstack{$\kappa = 3.17$, $\omega = 0.10$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.13$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$,  $\phi_{2G} = 0.19$,\\ $\phi_{3, A} = 0.35$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
         GY94 + $\Gamma\omega$ & {\shortstack{$\alpha_\omega = 0.51$, $\beta_\omega = 3.92$, $\kappa = 3.49$, \\$\phi_{1,A} = 0.32$, $\phi_{1,C} = 0.14$, $\phi_{1,G} = 0.28$, \\$\phi_{2,A} = 0.38$, $\phi_{2,C} = 0.18$, $\phi_{2,G} = 0.20$, \\$\phi_{3,A} = 0.355$, $\phi_{3,C} = 0.19$, $\phi_{3,G} = 0.21$}}\\ \hline
         ExpCM(H1) &  {\shortstack{$\beta = 1.22$, $\kappa = 3.24$, $\omega = 0.13$,\\ $\phi_A = 0.37$, $\phi_C = 0.18$, $\phi_G = 0.23$}}\\ \hline
      \end{tabular}
\end{supptable}
\clearpage

@@

\skhcomment{Notes from Sarah to Jesse}

\skhcomment{(1) Within subtype amino acid identity. I didn't want to do a brute force approach so I took 1 million samples the amino-acid identity between pairs of sequences and took the minimum identity of these samples within a subtype. Do you think this is sufficient?}

\skhcomment{(2) The asymptotic sequence divergence of the GY94 model. Currently, I have a value of 0.059 for the asymptotic divergence. I got this value by simply putting a very large value of t into the formula. I tried to do out the math and find the \textit{actual} limit but I got a value 0.0193. I will attach my notes and maybe there is an obvious place where the formula is wrong...}

\begin{equation}
\label{eq:limit}
\end{equation}
